---
title: "Code lists for anxiety and depression psychological therapies"
format: 
  html:
    code-hidden: true
editor: visual
---

### Background

This document compiles the codelist of psychological therapies for Depression and Anxiety in the UK General practice. The codelist benefits from the NICE guideline recommendations on the management of anxiety and depression. Specific documents reviewed were:

-   [Depression in adults: treatment and management](https://www.nice.org.uk/guidance/ng222/chapter/Recommendations)
-   [Generalised anxiety disorder and panic disorder in adults: management](https://www.nice.org.uk/guidance/cg113/chapter/Recommendations)

Furthermore, the NHSD codelists for anxiety and depression available on OpenCodelist were adapted. These include:

-   Depression support and treatment codes

-   Anxiety support and treatment codes

### Methodology

A two-step approach was taken for the development of this codelist:

**Step 1: Desk reviews of relevant guidelines and existing codelists for anxiety and depression**

This codelist benefits from the NICE guideline recommendations on the management of anxiety and depression. Specific documents reviewed were:

1.  **Depression in adults: treatment and management**

2.  **Generalised anxiety disorder and panic disorder in adults: management**

The recommended therapies outlined in the guidelines and existing NHSD codelists for anxiety and depression, available on OpenCodelist, were used in developing a draft codelist. The adapted codelists were:

1.  Depression support and treatment codes

2.  Anxiety support and treatment codes

**Step 2: A methodology-informed session with a General Practitioner** A methodology-informed session was held with a General Practitioner to understand the approach to coding psychological therapies in eHR. This informed the development of improved search terms and inclusion/exclusion criteria.

**Inclusion and exclusion criteria**

This codelist is designed to maximize sensitivity by including all psychological therapy-related codes potentially applicable to anxiety and depression, regardless of explicit indication. Specifically,

-   Psychological therapies codes specific to anxiety and depression were included

-   Psychological therapies codes for conditions related to anxiety and depression (e.g Insomnia, bipolar disorder, psychosis) were included

-   Generic psychological therapies codes (e.g. Cognitive Behavioural Therapy) not specific to other conditions were included

-   Common codes (based on GP's perspective) that indicated receipt of psychological therapies were included (e.g. seen by psychologists)

-   Psychological therapies codes specific to other conditions not related to anxiety and depression were excluded 

#### Search Terms

-   Guided self-help

-   Digital cognitive behavioural therapy for depression

-   Cognitive Behavioural Therapy

-   Behavioural Activation

-   Group Exercise

-   Group mindfulness and meditation

-   Interpersonal psychotherapy

-   Counselling

-   Short-term psychodynamic psychotherapy

-   Individual problem solving

-   Behavioural couple therapy

-   Anxiety

-   Depression

-   Individual guided self-help

-   Individual non-facilitated self-help

-   Psycho-education

-   Education about anxiety

-   Counselling for anxiety

-   Referral to improving access to psychological therapies program

-   Talking therapies

### Results: Code Lists Usage Patterns 

```{r}
#| label: Load libraries
#| message: false
#| warning: false
#| include: false


#Load relevant libraries 
library(opencodecounts) 
library(tidyverse)
library(scales)
library(gt)
```

Load codelist: <https://www.opencodelists.org/codelist/user/Kunle_Oreagba/psychological-therapies-coding-for-anxiety-and-depression/2d533ab8/> from opencodelist

```{r}
#| label: Load codelist
#| echo: true
#| message: false
#| warning: false


#Load codelists from opencodelist

depr_anx_cod <- get_codelist("user/Kunle_Oreagba/psychological-therapies-coding-for-anxiety-and-depression/2d533ab8")

```

##### A. Summary of depression and anxiety psychological therapies code usage 

```{r}
#| label: Table of Summary of Codelist Usage
#| echo: false
#| message: false
#| warning: false

#Generate summary of the codelists and usage 
codelist_summary <- snomed_usage %>%
    filter(snomed_code %in% depr_anx_cod$code) %>%
    group_by(snomed_code, description) %>%
    summarise(usage = sum(usage), .groups = "drop") %>%
    mutate(percent_usage = usage / sum(usage)) %>%
    arrange(desc(usage))

#Generate a table of code usage  
codelist_summary |>
  gt() |>
  cols_align(
  align = "left",
  columns = c(snomed_code, description))|>
  fmt_percent(
    columns = c(percent_usage),
    decimals = 2
  ) |>
    tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(columns = c(snomed_code, description, usage, percent_usage))
  ) |>
  tab_header(
    title = md("**Overview of Depression and Anxiety Code Usage**")
  ) |>
  tab_options(
    # Title font size
    heading.title.font.size = px(16))|>
  tab_style(
    style = cell_text(align = "left"),
    locations = cells_title(groups = "title")
  )

    
```

##### Top 15 most utilized depression and anxiety psychological therapies codes

```{r}
#| label: Top 15 most utilized codes
#| echo: false
#| message: false
#| warning: false

#Slice the top 15 code usage 
top_codelist <- codelist_summary %>%
  arrange(desc(usage)) %>%
  slice_head (n = 15) %>%
  mutate(description = factor(description, levels = rev(description)))  # Fix order for horizontal bars

# Plot a horizontal bar chart


# Calculate dynamic upper limit
y_max <- max(top_codelist$usage) + 100000

ggplot(top_codelist, aes(x = description, y = usage)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  geom_text(aes(label = paste0(round(usage / 1000), "k")), hjust = -0.1, size = 2)+
  labs(
    title = "Top 15 most used codes for depression and anxiety psychological therapies",
    x = "Therapies",
    y = "Usage"
  ) +
  scale_y_continuous(
    limits = c(0, y_max),
    labels = function(x) paste0(x / 1000, "k")) +
  theme_minimal()+
  theme(
    plot.margin = margin(t = 20, r = 10, b = 10, l = 10),
    plot.title = element_text(hjust = 1.1, size = 12, face = "bold")
  )

```

```{r}
# Load packages
#| echo: false


dep_anx_tx_cod_usage <- snomed_usage |>
    filter(snomed_code %in% depr_anx_cod$code)

#Extract IAPT code usage 

iapt_codes_usage <- dep_anx_tx_cod_usage |> 
  filter(str_detect(str_to_lower(description), "access to psychological therapies programme"))


# Extract semantic tag and add new variable
cod_usage_semantic_tag <- iapt_codes_usage |> 
  mutate(
    description_short = str_remove(description, " \\([^()]+\\)$"),
    semantic_tag = str_extract(description, "\\(([^()]+)\\)$"),
    semantic_tag = str_replace_all(semantic_tag, "[()]", "")
  )

#Group by semantic tag
group_code_semantic <- cod_usage_semantic_tag |> 
  group_by(snomed_code, description_short, semantic_tag) |> 
  summarise(usage = sum(usage), .groups = "drop") |>
  mutate(ratio_usage = usage / sum(usage)) |>
  arrange(-usage)

# Create table grouped by semantic_tag
group_code_semantic |> 
  group_by(semantic_tag) |> 
  gt() |> 
  fmt_percent(
    columns = c(ratio_usage),
    decimals = 2
  ) |> 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = semantic_tag)
  ) |>
  tab_header(
    title = md("**Code Usage by Semantic Tag**")
  )


```

#### Plot of code usage by semantic tag

```{r}
#| echo: false
#| message: false
#| warning: false


library(scales)

# Create plot with facets using facet_wrap()
ggplot(cod_usage_semantic_tag, 
  aes(
    y = usage, 
    x = end_date,
    colour = description
    )) + 
  geom_line(alpha = 0.2) + 
  geom_point() +
  labs(
    title = "Yearly counts of improving access to psychological therapies programme",
    x = NULL,
    y = NULL,
    colour = NULL,
    caption = "Data from NHS England SNOMED Code Usage in Primary Care.") + 
  scale_x_date(
    date_breaks = "1 year",
    labels = label_date_short()) +
  scale_y_continuous(
    labels = label_number(),
    limits = c(0, NA)
    ) +
  facet_wrap(~semantic_tag, scales = "free_y") +
  theme_bw() +
  theme(legend.position = "bottom",
        plot.margin = margin(t = 20, r = 20, b = 20, l = 20)) +
  guides(colour = guide_legend(ncol = 2, byrow = TRUE))
```
