---
title: "Codelist development for and anxiety disorders and depression diagnoses"
toc: true
toc-depth: 2
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    other-links:
      - text: opencodecounts R pkg
        href: https://bennettoxford.github.io/opencodecounts/
      - text: SNOMED Code Usage
        href: https://digital.nhs.uk/data-and-information/publications/statistical/mi-snomed-code-usage-in-primary-care
execute:
  cache: true
editor: source
---

# Load packages

```{r}
#| label: setup
#| message: false
#| results: 'hide'
#| warning: false
#| echo: true

library(opencodecounts) 
library(tidyverse)
library(scales)
library(gt)
library(here)

source(here("lib", "fun_tables.R"))
source(here("lib", "fun_data.R"))
source(here("lib", "fun_figures.R"))
```

# NHS Primary Care Refsets

Load [Primary Care Domain Refset](https://digital.nhs.uk/data-and-information/data-collections-and-data-sets/data-collections/quality-and-outcomes-framework-qof/quality-and-outcome-framework-qof-business-rules/primary-care-domain-reference-set-portal) codelists from [OpenCodelists](https://www.opencodelists.org/codelist/nhsd-primary-care-domain-refsets/) for anxiety diagnostic codes ([`anx_cod`](https://www.opencodelists.org/codelist/nhsd-primary-care-domain-refsets/anxiety-diagnostic-codes/20241205/)) and depression diagnostic codes ([`depr_cod`](https://www.opencodelists.org/codelist/nhsd-primary-care-domain-refsets/depr_cod/20241205/)). 
We search all SNOMED CT descriptions for terms that we want to review for possible inclusion in the existing codelists (e.g., "depression", "anxiety").


```{r}
#| label: load-codelists

depr_cod <- get_codelist(
  "nhsd-primary-care-domain-refsets/depr_cod/20241205"
)

anx_cod <- get_codelist(
  "nhsd-primary-care-domain-refsets/anxiety-diagnostic-codes/20241205"
)

# Remove all codes that are in our codelist
snomed_usage_without_depr_cod <- snomed_usage |> 
  filter(!snomed_code %in% depr_cod$code)

# Search for related codes (e.g., with "depression" in description) and review
snomed_usage_without_depr_cod_review <- snomed_usage_without_depr_cod |> 
  filter(str_detect(description, "depression"))

# Remove all codes that are in our codelist
snomed_usage_without_anx_cod <- snomed_usage |> 
  filter(!snomed_code %in% anx_cod$code)

# Search for related codes (e.g., with "depression" in description) and review
snomed_usage_without_anx_cod_review <- snomed_usage_without_anx_cod |> 
  filter(str_detect(description, "anxiety"))
```

## Total code usage

::: {.panel-tabset}

### Depression

The below table summarises the frequency of each code from 2011/12 to 2023/24 included in the depression codelist.
There were *n* = `r nrow(snomed_usage_without_depr_cod_review)` codes with the term `"depression"` in their description that are not in our codelist.
These codes could be reviewed by a clinical expert to determine if they should be included here.
For now, we will assume that they are not relevant as they have not been included in the NHSD refset.


```{r}
#| label: tab-depr-cod-counts
#| message: false
#| warning: false

depr_cod_counts <- snomed_usage |> 
  filter(snomed_code %in% depr_cod$code) |> 
  get_semantic_tag() |> 
  summarise_code_counts() |> 
  slice_max(n = 20, order_by = usage)

# Display formatted table
tab_depr_cod_counts <- depr_cod_counts |>
  create_table_sem_tag(
    title = md("Top 15 depression diagnosis codes (`depr_cod`)"),
    subtitle = "Timeframe: August 2011 to July 2024"
    )

gtsave(
  tab_depr_cod_counts,
  here("output", "tables", "cod_devel", "dx", "tab_depr_cod_usage.png")
  )

tab_depr_cod_counts
```

### Anxiety disorders

The below table summarises the frequency of each code from 2011/12 to 2023/24 included in the anxiety codelist.
There were *n* = `r nrow(snomed_usage_without_anx_cod_review)` codes with the term `"anxiety"` in their description that are not in our codelist. These codes could be reviewed by a clinical expert to determine if they should be included here. For now, we will assume that they are not relevant as they have not been included in the NHSD refset.


```{r}
#| label: tab-anx-cod-counts
#| message: false
#| warning: false

anx_cod_counts <- snomed_usage |> 
  filter(snomed_code %in% anx_cod$code) |> 
  get_semantic_tag() |> 
  summarise_code_counts() |> 
  slice_max(n = 20, order_by = usage)

# Display formatted table
tab_anx_cod_counts <- anx_cod_counts |>
  create_table_sem_tag(
    title = md("Top 15 anxiety diagnosis codes (`anx_cod`)"),
    subtitle = "Timeframe: August 2011 to July 2024"
    )

gtsave(
  tab_anx_cod_counts,
  here("output", "tables", "cod_devel", "dx", "tab_anx_cod_usage.png")
  )

tab_anx_cod_counts
```

:::

## Trends over time

The below figure shows the coding of anxiety disorders and depression diagnoses, as specified in the NHS Primary Care Domain Refsets ([anx_cod](https://www.opencodelists.org/codelist/nhsd-primary-care-domain-refsets/anxiety-diagnostic-codes/20241205/), [depr_cod](https://www.opencodelists.org/codelist/nhsd-primary-care-domain-refsets/depr_cod/20241205/)).


```{r}
#| label: fig-anx-dep-cod-trends
#| fig-cap: "Yearly counts of anxiety disorder and depression diagnoistic codes."
#| fig-height: 4.5
#| fig-width: 8
#| warning: false

anx_cod_depr_cod_usage <- snomed_usage |> 
  mutate(
    depr_cod = snomed_code %in% depr_cod$code,
    anx_cod = snomed_code %in% anx_cod$code
    ) |> 
  pivot_longer(
    cols = c(depr_cod, anx_cod),
    names_to = "codelist") |> 
  filter(value == TRUE) |> 
  group_by(end_date, codelist) |> 
  summarise(count = sum(usage, na.rm = TRUE))

fig_depr_anx_cod_trends <- anx_cod_depr_cod_usage |> 
  mutate(codelist = factor(
    codelist,
    levels = c("depr_cod", "anx_cod"), 
    labels = c("Depression diagnostic codes", "Anxiety diagnostic codes"))) |> 
  ggplot(
      aes(
        y = count,
        x = end_date,
        colour = codelist,
        shape = codelist
      )
    ) +
    geom_line(alpha = 0.2) +
    geom_point() +
    labs(
      title = "Yearly counts of anxiety disorder and depression diagnoistic codes",
      x = NULL,
      y = NULL,
      colour = NULL,
      shape = NULL,
      caption = "Data from NHS England SNOMED Code Usage in Primary Care."
    ) +
    scale_x_date(
      date_breaks = "1 year",
      labels = label_date_short()
    ) +
    scale_y_continuous(
      labels = label_number(accuracy = 1),
      limits = c(0, NA)
    ) +
    theme_bw() +
    theme(legend.position = "bottom") +
    scale_colour_viridis_d(end = .8)

ggsave(
  filename = here("output", "figures", "cod_devel", "dx", "fig_depr_anx_cod_trends.png"),
  fig_depr_anx_cod_trends,
  height = 4,
  width = 8
)

fig_depr_anx_cod_trends
```
