---
title: "Codelist development for psychological therapies for anxiety disorders and depression "
toc: true
toc-depth: 2
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    other-links:
      - text: opencodecounts R pkg
        href: https://bennettoxford.github.io/opencodecounts/
      - text: SNOMED Code Usage
        href: https://digital.nhs.uk/data-and-information/publications/statistical/mi-snomed-code-usage-in-primary-care
editor: source
---

```{r}
#| label: setup
#| message: false
#| warning: false
#| include: false

# Load relevant libraries 
library(opencodecounts) 
library(here)
library(tidyverse)
library(scales)
library(gt)
library(scales)
library(stringr)

source(here("lib", "fun_tables.R"))
source(here("lib", "fun_data.R"))
source(here("lib", "fun_figures.R"))
```

# Background

This document compiles the codelist of psychological therapies for Depression and Anxiety in the UK General practice. The codelist benefits from the NICE guideline recommendations on the management of anxiety and depression. Specific documents reviewed were:

- [NICE Guidance: Depression in adults: treatment and management](https://www.nice.org.uk/guidance/ng222/chapter/Recommendations)
- [NICE Guidance: Generalised anxiety disorder and panic disorder in adults: management](https://www.nice.org.uk/guidance/cg113/chapter/Recommendations)

# Methods

## Codelist development

### Codelist development

A two-step approach was taken for the development of this codelist:

1.  Review of relevant guidelines and existing codelists for anxiety and depression
    - Depression support and treatment codes: [`depsupp_cod`](https://www.opencodelists.org/codelist/nhsd-primary-care-domain-refsets/depsupp_cod/20250627/)
    - Anxiety support and treatment codes: [`anxsupp_cod`](https://www.opencodelists.org/codelist/nhsd-primary-care-domain-refsets/anxsupp_cod/20250627/)
2. Review of initial codelists with a General Practitioner. The aim was to understand the approach to coding psychological therapies by GPs in EHR. This further informed the development of improved search terms and inclusion/exclusion criteria.

### Inclusion and exclusion

This codelist is designed to maximize sensitivity by including all psychological therapy-related codes potentially applicable to anxiety and depression, regardless of explicit indication, specifically:

- Psychological therapies codes specific to anxiety and depression were included
- Psychological therapies codes for conditions related to anxiety and depression (e.g Insomnia) were included
- Generic psychological therapies codes (e.g. Cognitive Behavioural Therapy) not specific to other conditions were included
- Common codes (based on GP's perspective) that indicated receipt of psychological therapies were included (e.g. seen by psychologists)
- Psychological therapies codes specific to other conditions not related to anxiety and depression were excluded 

### Search terms

Following the review of relevant guidelines and discussions with a GP the following search terms were used to review the description of SNOMED CT codes:

- Guided self-help
- Digital cognitive behavioural therapy for depression
- Cognitive Behavioural Therapy
- Behavioural Activation
- Group Exercise
- Group mindfulness and meditation
- Interpersonal psychotherapy
- Counselling
- Short-term psychodynamic psychotherapy
- Individual problem solving
- Behavioural couple therapy
- Anxiety
- Depression
- Individual guided self-help
- Individual non-facilitated self-help
- Psycho-education
- Education about anxiety
- Counselling for anxiety
- Referral to improving access to psychological therapies program
- Talking therapies

# Results

## Codelist for psychological therapies

We developed a codelist including psychological therapies for anxiety disorders and depression.

```{r}
#| label: load-codelists
#| echo: true
#| message: false
#| warning: false

# Load codelists from opencodelist
depr_anx_psy_tx_cod <- get_codelist(
  "user/Kunle_Oreagba/psychological-therapies-coding-for-anxiety-and-depression/3f6fc6e3"
  )
```

## Total code usage

```{r}
#| label: tab-summary-codelist-usage
#| echo: false
#| message: false
#| warning: false

# Generate summary of the codelists and usage 
codelist_summary <- snomed_usage %>%
    filter(snomed_code %in% depr_anx_psy_tx_cod$code) %>%
    group_by(snomed_code, description) %>%
    summarise(usage = sum(usage), .groups = "drop") %>%
    mutate(percent_usage = usage / sum(usage)) %>%
    arrange(desc(usage))

# Generate a table of code usage  
all_psych_tx_counts <- snomed_usage |>
  filter(snomed_code %in% depr_anx_psy_tx_cod$code) |> 
  get_semantic_tag() |> 
  summarise_code_counts()

# Display formatted table
tab_all_psych_tx <- all_psych_tx_counts |>
  create_table_sem_tag(
    title = "Summary of selected codes for psychological therapies",
    subtitle = "Timeframe: August 2011 to July 2024"
    )

gtsave(
  tab_all_psych_tx,
  here("output", "tables", "tab_all_psych_tx_sem_tag_usage.png")
  )

tab_all_psych_tx
```

### Most used codes

```{r}
#| label: top15-codes
#| echo: false
#| message: false
#| warning: false

# Slice the top 15 code usage 
top_codelist <- codelist_summary %>%
  arrange(desc(usage)) %>%
  slice_head (n = 15) %>%
  mutate(description = factor(description, levels = rev(description)))  # Fix order for horizontal bars

# Plot a horizontal bar chart

# Calculate dynamic upper limit
y_max <- max(top_codelist$usage) + 100000

ggplot(top_codelist, aes(x = description, y = usage)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  geom_text(aes(label = paste0(round(usage / 1000), "k")), hjust = -0.1, size = 2)+
  labs(
    title = "Top 15 most used codes for depression and anxiety psychological therapies",
    x = "Therapies",
    y = "Usage"
  ) +
  scale_y_continuous(
    limits = c(0, y_max),
    labels = function(x) paste0(x / 1000, "k")) +
  theme_minimal()+
  theme(
    plot.margin = margin(t = 20, r = 10, b = 10, l = 10),
    plot.title = element_text(hjust = 1.1, size = 12, face = "bold")
  )

# Obtain total IAPT code usage to be used for interpretation 
iapt_total = codelist_summary$usage[codelist_summary$description == "Referral to improving access to psychological therapies programme (procedure)"] + codelist_summary$usage[codelist_summary$description == "Self-referral to Improving Access to Psychological Therapies programme (procedure)"]


# Obtain total "Seen by Psychologist code"
seen_by_psychologist = codelist_summary$usage[codelist_summary$description == "Seen by psychologist (finding)"]

# Obtain total acceted for talking therapy code 
talking_therapy = codelist_summary$usage[codelist_summary$description == "Accepted by psychological talking therapy service (finding)"]
```

Top 15 most recorded psychological therapy codes in UK primary care had a total of `r round(sum(top_codelist$usage)/1000000, 2)` million recordings between 2012 and 2024. **Referral/Self-referral to Improving access to psychological therapies** was the most recorded code with over `r round (iapt_total/1000000, 2)` million recordings accounting for `r round((iapt_total/sum(codelist_summary$usage))*100, 1)` %, followed by acceptance for psychological talking therapy service with `r  round(talking_therapy/1000, 0)` thousand recordings. There were over `r round(seen_by_psychologist/1000, 0)` thousand recordings as **Seen by Psychologists.**

## NHS Talking Therapies (NHS TT)

Previously known as IAPT.

```{r}
#| label: semantic-tags-for-iapt-codes
#| echo: false
#| message: false
#| warning: false

# Extract the codelist from SNOMED
codelist <- snomed_usage |>
    filter(snomed_code %in% depr_anx_psy_tx_cod$code)

# Extract IAPT and talking therapies codes 

# Set search terms 
search_term1 = "access to psychological therapies programme"
search_term2 = "psychological talking therapy"
search_terms = c(search_term1,search_term2)

iapt_codes_usage <- codelist |> 
  filter(str_detect(str_to_lower(description), str_c(str_to_lower(search_terms), collapse = "|")))


# Extract semantic tag and add new variable
iapt_usage_semantic_tag <- iapt_codes_usage |> 
  mutate(
    description_short = str_remove(description, " \\([^()]+\\)$"),
    semantic_tag = str_extract(description, "\\(([^()]+)\\)$"),
    semantic_tag = str_replace_all(semantic_tag, "[()]", "")
  )

# Group by semantic tag
group_iapt_semantic <- iapt_usage_semantic_tag |> 
  group_by(snomed_code, description_short, semantic_tag) |> 
  summarise(usage = sum(usage), .groups = "drop") |>
  mutate(percentage = usage / sum(usage)) |>
  arrange(-usage)

# Create table grouped by semantic_tag
group_iapt_semantic |> 
  group_by(semantic_tag) |> 
  gt() |> 
  cols_align(
  align = "left",
  columns = c(snomed_code, description_short))|>
  fmt_percent(
    columns = c(percentage),
    decimals = 2
  ) |> 
    tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(columns = c(snomed_code, description_short, usage, percentage))
  )|>
  tab_header(
    title = md("**IAPT code usage by clinical concepts**")
  )|>
    tab_options(
    # Title font size
    heading.title.font.size = px(16))|>
  tab_style(
    style = cell_text(align = "left"),
    locations = cells_title(groups = "title")
  )
```

```{r}
#| label: fig-iapt-trends
#| fig-cap: "Yearly counts of codes with the term 'access to psychological therapies programme' or 'psychological talking therapy' in their description from 2011/12 to 2023/24."
#| fig-height: 6
#| fig-width: 8
#| warning: false

# Create plot with facets using facet_wrap()
fig_psy_tx_iapt_trends <- plot_trajectory(
  iapt_usage_semantic_tag,
  ncol_legend = 1)

ggsave(
  filename = here("output", "figures", "fig_psy_tx_iapt_trends.png"),
  fig_psy_tx_iapt_trends,
  height = 6,
  width = 8
)

fig_psy_tx_iapt_trends
```

There has been an upward trend in self-referrals to Improving Access to Psychological Therapies (IAPT) since the end of the COVID-19 pandemic(2021). Similarly, the acceptance to psychological talking therapy has seen an upward trend since 2019, with a sharp increase in 2024 (84.5%) as compared to 2023.

## Cognitive Behavioural Therapy (CBT)

```{r}
#| label: cbt-patterns
#| echo: false
#| message: false
#| warning: false

# Check the Overview of all Cognitive Behavioural Therapy Codes 

#Set search terms 
term1 = "cognitive behavioural therapy"
term2 = "cognitive behavioral therapy"
term3 = "cognitive and behavioral therapy"
term4 = "cognitive and behavioural therapy"
term5 = "cognitive therapy"
search_terms = c(term1, term2, term3, term4, term5)

# Extract CTB related codes
cbt_codes_usage <- codelist |> 
 filter(str_detect(str_to_lower(description), str_c(str_to_lower(search_terms), collapse = "|")))

cbt_codes_summary <- cbt_codes_usage |>
  group_by(snomed_code, description) %>%
  summarise(usage = sum(usage), .groups = "drop") %>%
  mutate(percent_usage = usage / sum(usage)) %>%
  arrange(desc(usage))

# Generate Table of Summary for CBT Codes 
cbt_codes_summary |>
  gt() |>
  cols_align(
  align = "left",
  columns = c(snomed_code, description))|>
  fmt_percent(
    columns = c(percent_usage),
    decimals = 2
  ) |>
    tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(columns = c(snomed_code, description, usage, percent_usage))
  ) |>
  tab_header(
    title = md("**Overview of Cognitive Behavioural Therapy Code Usage**")
  ) |>
  tab_options(
    # Title font size
    heading.title.font.size = px(16))|>
  tab_style(
    style = cell_text(align = "left"),
    locations = cells_title(groups = "title")
  )
```

### Top 10 CBT codes

```{r}
#| label: top-cbt-codes
#| echo: false
#| message: false
#| warning: false

# Slice the top 10 CBT code usage 
top_cbt <- cbt_codes_summary %>%
  arrange(desc(usage)) %>%
  slice_head (n = 10) %>%
  mutate(description = factor(description, levels = rev(description)))  # Fix order for horizontal bars

# Plot a horizontal bar chart
# Calculate dynamic upper limit
max_value <- max(top_cbt$usage) + 10000

ggplot(top_cbt, aes(x = description, y = usage)) +
  geom_bar(stat = "identity", fill = "orange") +
  coord_flip() +
  geom_text(aes(label = paste0(round(usage / 1000), "k")), hjust = -0.1, size = 2)+
  labs(
    title = "Top 10 most used cognitive behavioural therapy codes",
    x = "Therapies",
    y = "Usage"
  ) +
  scale_y_continuous(
    limits = c(0, max_value),
    labels = function(x) paste0(x / 1000, "k")) +
  theme_minimal()+
  theme(
    plot.margin = margin(t = 20, r = 10, b = 10, l = 10),
    plot.title = element_text(hjust = 1.1, size = 12, face = "bold")
  )
```

24 disease and non-disease-specific Cognitive Behavioural Therapy codes were identified. Referral for Cognitive Behavioural Therapy was the most recorded event with `r round (cbt_codes_summary$usage[cbt_codes_summary$snomed_code == 519101000000109] / 1000, 0)` thousand recordings, followed by by Cognitive and behavioural therapy with `r round(cbt_codes_summary$usage[cbt_codes_summary$snomed_code == 228557008] / 1000, 0)` thousand recordings. Guided self-help CBT had `r round(cbt_codes_summary$usage[cbt_codes_summary$snomed_code == 444175001] / 1000, 0)` thousand recordings, a similar frequency with Computerized CBT (`r round(cbt_codes_summary$usage[cbt_codes_summary$snomed_code == 429329005] / 1000, 0)` thousand recordings). Fewer number of codings were disease specific. For instance, CBT for depression had only `r round(cbt_codes_summary$usage[cbt_codes_summary$snomed_code == 149631000000108] / 1000, 0)` thousand recordings.

### Pattern of CBT codes usage by clinical concepts (only top 10 codes)

```{r}
#| label: cbt-semantic-tag
#| echo: false
#| message: false
#| warning: false

# Extract semantic tag and add new variable
cbt_usage_semantic_tag <- cbt_codes_usage |> 
  mutate(
    description_short = str_remove(description, " \\([^()]+\\)$"),
    semantic_tag = str_extract(description, "\\(([^()]+)\\)$"),
    semantic_tag = str_replace_all(semantic_tag, "[()]", "")
  )

#Group by semantic tag
group_cbt_semantic <- cbt_usage_semantic_tag |> 
  group_by(snomed_code, description_short, semantic_tag) |> 
  summarise(usage = sum(usage), .groups = "drop") |>
  mutate(percentage = usage / sum(usage)) |>
  arrange(-usage)

# Create table grouped by semantic_tag
group_cbt_semantic |> 
  group_by(semantic_tag) |> 
  gt() |> 
  cols_align(
  align = "left",
  columns = c(snomed_code, description_short))|>
  fmt_percent(
    columns = c(percentage),
    decimals = 2
  ) |> 
    tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(columns = c(snomed_code, description_short, usage, percentage))
  )|>
  tab_header(
    title = md("**CBT code usage by clinical concepts**")
  )|>
    tab_options(
    # Title font size
    heading.title.font.size = px(16))|>
  tab_style(
    style = cell_text(align = "left"),
    locations = cells_title(groups = "title")
  )
```

```{r}
#| label: top10-cbt-codes-by-sem-tag
#| echo: false
#| message: false
#| warning: false

# Extract the first 5 most used CBT codes 
search_first_four = cbt_codes_summary$snomed_code[1:5]

# Filter semantic tags for the top codes 
top_cbt_semantic_tags <- cbt_usage_semantic_tag |>
  filter(snomed_code %in% search_first_four)

# Wrap description to enable full display in the chart
top_cbt_semantic_tags$description <- str_wrap(top_cbt_semantic_tags$description, width = 60)

# Create plot with facets using facet_wrap()
ggplot(top_cbt_semantic_tags, 
  aes(
    y = usage, 
    x = end_date,
    colour = description
    )) + 
  geom_line(alpha = 0.2) + 
  geom_point() +
  labs(
    title = 'Yearly Usage of Cognitive Behavioural Therapy Codes',
    x = 'Year',
    y = 'Usage',
    colour = NULL,
    caption = "Data from NHS England SNOMED Code Usage in Primary Care.") + 
  scale_x_date(
    date_breaks = "2 years",
    labels = label_date_short()) +
  scale_y_continuous(
    labels = label_number(),
    limits = c(0, NA)
    ) +
  facet_wrap(~semantic_tag, scales = "free_y") +
  theme_bw() +
  theme(legend.position = "bottom",
        plot.margin = margin(t = 20, r = 40, b = 10, l = 10)) +
  guides(colour = guide_legend(ncol = 2, byrow = TRUE))
```

Recordings on referrals for CBT declined between 2016 and 2021, but have since seen an increasing trend. There was a sharp increased recording of computerized CBT in 2020 compared to other years. There has been a rapid increase in CBT (therapy/regime) recordings since 2022. Recordings oon guided self-help CBT steadied between 2012 and 2020, with a slight increase since then.

## "Seen by Psychologist"

```{r}
#| label: seen-by-psych-sem-tag
#| echo: false
#| message: false
#| warning: false

# Extract Seen by Psychologists codes  
# Set search terms 
key_word = "psychologist"

seen_by_psy_codes_usage <- codelist |> 
 filter(str_detect(str_to_lower(description), str_c(str_to_lower(key_word), collapse = "|")))


# Extract semantic tag and add new variable
seen_by_psy_semantic_tag <- seen_by_psy_codes_usage |> 
  mutate(
    description_short = str_remove(description, " \\([^()]+\\)$"),
    semantic_tag = str_extract(description, "\\(([^()]+)\\)$"),
    semantic_tag = str_replace_all(semantic_tag, "[()]", "")
  )


#Generate a table for seen by psychologist clinical concepts code usage 
group_seen_psy_semantic <- seen_by_psy_semantic_tag |> 
  group_by(snomed_code, description_short, semantic_tag) |> 
  summarise(usage = sum(usage), .groups = "drop") |>
  mutate(percentage = usage / sum(usage)) |>
  arrange(-usage)

# Create table grouped by semantic_tag
group_seen_psy_semantic |> 
  group_by(semantic_tag) |> 
  gt() |> 
  cols_align(
  align = "left",
  columns = c(snomed_code, description_short))|>
  fmt_percent(
    columns = c(percentage),
    decimals = 2
  ) |> 
    tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(columns = c(snomed_code, description_short, usage, percentage))
  )|>
  tab_header(
    title = md("**Seen by Psychologists Codes by clinical concepts**")
  )|>
    tab_options(
    # Title font size
    heading.title.font.size = px(16))|>
  tab_style(
    style = cell_text(align = "left"),
    locations = cells_title(groups = "title")
  )
```

```{r}
#| label: fig-seen-by-psy
#| echo: false
#| message: false
#| warning: false

# Wrap description to enable full display in the chart
seen_by_psy_semantic_tag$description <- str_wrap(seen_by_psy_semantic_tag$description, width = 60)

# Create plot with facets using facet_wrap()
ggplot(seen_by_psy_semantic_tag, 
  aes(
    y = usage, 
    x = end_date,
    colour = description
    )) + 
  geom_line(alpha = 0.2) + 
  geom_point() +
  labs(
    title = 'Yearly Usage of Seen by Psychologist Codes',
    x = 'Year',
    y = 'Usage',
    colour = NULL,
    caption = "Data from NHS England SNOMED Code Usage in Primary Care.") + 
  scale_x_date(
    date_breaks = "2 years",
    labels = label_date_short()) +
  scale_y_continuous(
    labels = label_number(),
    limits = c(0, NA)
    ) +
  facet_wrap(~semantic_tag, scales = "free_y") +
  theme_bw() +
  theme(legend.position = "bottom",
        plot.margin = margin(t = 20, r = 40, b = 10, l = 10)) +
  guides(colour = guide_legend(ncol = 2, byrow = TRUE))
```
