---
title: "Analysis of anxiety disorders and depression psychological therapies codes"
toc: true
toc-depth: 2
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    other-links:
      - text: opencodecounts R pkg
        href: https://bennettoxford.github.io/opencodecounts/
      - text: SNOMED Code Usage
        href: https://digital.nhs.uk/data-and-information/publications/statistical/mi-snomed-code-usage-in-primary-care
editor: source
---

## Background 

This document describes the coding patterns for psychological therapies for anxiety disorders and depression in English primary care EHR. This analysis utilized a codelist developed from NHS England’s annual SNOMED CT Code Usage in Primary Care datasets, covering the period from 1st August 2011 to 31st July 2024.See below relevant references:

- [SNOMED Usage in Primary Care:](https://digital.nhs.uk/data-and-information/publications/statistical/mi-snomed-code-usage-in-primary-care)
- [Codelists:](https://www.opencodelists.org/codelist/user/Kunle_Oreagba/psychological-therapies-coding-for-anxiety-and-depression/415bd34a/)
- [Opencodecounts:](https://github.com/bennettoxford/opencodecounts)

Data was analysed primarily using the opencodecounts R package. 

```{r}
#| label: setup
#| message: false
#| warning: false
#| include: false

# Load relevant libraries
library(opencodecounts)
library(here)
library(tidyverse)
library(scales)
library(gt)
library(scales)
library(stringr)

source(here("lib", "fun_tables.R"))
source(here("lib", "fun_data.R"))
source(here("lib", "fun_figures.R"))

#Set search terms 

search_by_referral_service <- list(
    referral = c(
    "referral",
    "self-referral",
    "signposting"
  ),
  
  service = c(
    "to improving access to psychological therapies",
    "psychological talking therapy",
    "counsel",
    "psychoeducation",
    "cognitive behavioural therapy",
    "cognitive behavioral therapy",
    "cognitive and behavioral therapy",
    "cognitive and behavioural therapy",
    "cognitive therapy",
    "seen by",
    "admission by",
    "liaison with",
    "therapy started",
    "psychological therapy",
    "electroconvulsive",
    "discharge by",
    "problem solving",
    "under care",
    "self-help",
    "education",
    "behavioral activation therapy",
    "psychodynamic psychotherapy",
    "interpersonal psychotherapy",
    "dynamic interpersonal therapy",
    "psychodynamic therapy",
    "transcranial magnetic",
    "mindfulness",
    "psychotherapy",
    "couple therapy",
    "behavioural activation",
    "applied relaxation"
    
  )
)

```

```{r}
#| label: load-codelists
#| echo: false
#| message: false
#| warning: false

# Load codelists from opencodelist
depr_anx_psy_tx_cod <- get_codelist(
  "user/Kunle_Oreagba/psychological-therapies-coding-for-anxiety-and-depression/6a9bc643"
  )

#Filter codes from the codelists 
dep_anx_psy_tx_usage = snomed_usage |> 
  filter(snomed_code %in% depr_anx_psy_tx_cod$code)
```

## Preliminary Objectives 

Preliminary objectives of the analysis include to:

- Describe the total usage of anxiety disorders and psychological therapy codes over the study period 
- Identify the most frequently used psychological therapy codes in English Primary care 
- Examine pattern of code usage by:
  - Referral or signposting vs actual delivery of therapy
  - High-intensity vs low-intensity psychological therapies
  - Therapy indication, including specific codes for anxiety disorders, depression and cross-cutting codes


## Results 
### Total code usage for anxiety disorders and depression psychological therapies

```{r}
#| label: total-code-usage
#| echo: false
#| message: false
#| warning: false

#Obtain total code usage
dep_anx_tx_total_cod_usage = dep_anx_psy_tx_usage |>
group_by(end_date) |> 
    summarise(usage = sum(usage), .groups = "drop") |>
    mutate(ratio_usage = usage / sum(usage)) |>
    arrange(-usage)

```


```{r}
#| label: fig-total-trends
#| echo: false
#| fig-cap: "Yearly counts of psychological therapy codes' from 2011/12 to 2023/24."
#| fig-height: 4
#| fig-width: 8
#| warning: false
 
fig_psy_tx_dep_anx_trends = dep_anx_tx_total_cod_usage |>
  ggplot(
      aes(
        y = usage,
        x = end_date,
      )
    ) +
    geom_line(alpha = 0.2) +
    geom_point() +
    labs(
      x = NULL,
      y = NULL,
      colour = NULL,
      caption = "Data from NHS England SNOMED Code Usage in Primary Care.",
      #title = "Yearly counts of psychological therapy codes used in primary care"
    ) +
    scale_x_date(
      date_breaks = "1 year",
      labels = label_date_short()
    ) +
    scale_y_continuous(
      labels = label_number(accuracy = 1),
      limits = c(0, NA)
    ) +
    theme_bw() +
    theme(legend.position = "bottom") +
  scale_color_viridis_d()
fig_psy_tx_dep_anx_trends

ggsave(
  filename = here("output", "figures", "fig_psy_tx_dep_anx_total_trends.png"),
  fig_psy_tx_dep_anx_trends,
  height = 4,
  width = 8
)

```


### Top 15 most frequently used codes
```{r}
#| label: tab-most-used-codes
#| echo: false
#| message: false
#| warning: false
top = 15
dep_anx_tx_top_codes = dep_anx_psy_tx_usage |>
group_by(snomed_code, description) |> 
    summarise(usage = sum(usage), .groups = "drop") |>
    mutate(ratio_usage = usage / sum(usage)) |>
    arrange(-usage) |>
  slice_head(n = top)


tab_dep_anx_tx_top_codes = dep_anx_tx_top_codes |>
  gt() |>
    fmt_percent(
      columns = c(ratio_usage),
      decimals = 1
    ) |>
    tab_style(
      style = list(
        cell_text(font = "Courier New")
      ),
      locations = cells_body(columns = snomed_code)
    ) |>
    cols_label(
      snomed_code = "SNOMED code",
      description = "Description",
      usage = "Usage",
      ratio_usage = "%"
    ) |>
    tab_style(
      style = cell_text(weight = "bold"),
      locations = cells_column_labels(columns = everything())
    ) |>
    cols_align(
      align = "left",
      columns = snomed_code
    ) |>
    tab_options(table.width = "100%") |>
    tab_source_note(
      source_note = md(
        "Data scource: NHS England SNOMED Usage in Primary Care."
      )
    )

gtsave(
  tab_dep_anx_tx_top_codes,
  here("output", "tables", "tab_dep_anx_tx_top_codes.png")
  )

tab_dep_anx_tx_top_codes

```


### Usage trend by referral/signposting vs actual service offering

```{r}
#| label: fig-trends_referral_service
#| echo: false
#| fig-cap: "Comparison of yealy counts of psychological therapy codes for referrals and actual service ' from 2011/12 to 2023/24."
#| fig-height: 4
#| fig-width: 8
#| warning: false
search_patterns_referrals <- map(search_by_referral_service, ~ str_c(.x, collapse = "|"))

psy_tx_referral_cat = dep_anx_psy_tx_usage |>
  filter_terms(unlist(search_by_referral_service))|>
  add_proms_match(search_patterns_referrals)|>
  
#Some description that contains word referral and actual service are categorized twice e.g. referral;service
  
#Addressing by replacing referral;service with referral
  
  mutate(proms = str_replace(proms, "referral;service", "referral"))|>
  group_by(end_date, proms) |> 
  summarise(usage = sum(usage), .groups = "drop") |>
  mutate(ratio_usage = usage / sum(usage)) |>
  arrange(end_date)

fig_psy_tx_service_referral_trends = psy_tx_referral_cat |>
  ggplot(
      aes(
        y = usage,
        x = end_date,
        colour = proms
      )
    ) +
    geom_line(alpha = 0.2) +
    geom_point() +
    labs(
      x = NULL,
      y = NULL,
      colour = NULL,
      caption = "Data from NHS England SNOMED Code Usage in Primary Care.",
      #title = "Yearly counts of psychological therapy codes used in primary care"
    ) +
    scale_x_date(
      date_breaks = "1 year",
      labels = label_date_short()
    ) +
    scale_y_continuous(
      labels = label_number(accuracy = 1),
      limits = c(0, NA)
    ) +
    theme_bw() +
    theme(legend.position = "bottom") +
  scale_color_viridis_d()
fig_psy_tx_service_referral_trends

ggsave(
  filename = here("output", "figures", "fig_psy_tx_service_referral_trends.png"),
  fig_psy_tx_service_referral_trends,
  height = 4,
  width = 8
)


```











