---
title: "Analysis of anxiety disorders and depression psychological therapies codes"
toc: true
toc-depth: 2
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    other-links:
      - text: opencodecounts R pkg
        href: https://bennettoxford.github.io/opencodecounts/
      - text: SNOMED Code Usage
        href: https://digital.nhs.uk/data-and-information/publications/statistical/mi-snomed-code-usage-in-primary-care
editor: source
---

# Background 

This document describes the coding patterns for psychological therapies for anxiety disorders and depression in English primary care EHR. This analysis utilized a codelist developed from NHS England’s annual SNOMED CT Code Usage in Primary Care datasets, covering the period from 1st August 2011 to 31st July 2024.See below relevant references:

- [SNOMED Usage in Primary Care:](https://digital.nhs.uk/data-and-information/publications/statistical/mi-snomed-code-usage-in-primary-care)
- [Codelists:](https://www.opencodelists.org/codelist/user/Kunle_Oreagba/psychological-therapies-coding-for-anxiety-and-depression/415bd34a/)
- [Opencodecounts:](https://github.com/bennettoxford/opencodecounts)

Data was analysed primarily using the opencodecounts R package. 

```{r}
#| label: setup
#| message: false
#| warning: false
#| include: false

# Load relevant libraries
library(opencodecounts)
library(here)
library(tidyverse)
library(scales)
library(gt)
library(scales)
library(stringr)

source(here("lib", "fun_tables.R"))
source(here("lib", "fun_data.R"))
source(here("lib", "fun_figures.R"))

#Set search terms 

```

```{r}
#| label: load-codelists
#| echo: true
#| message: false
#| warning: false

# Load codelists from opencodelist
depr_anx_psy_tx_cod <- get_codelist(
  "user/Kunle_Oreagba/psychological-therapies-coding-for-anxiety-and-depression/415bd34a"
  )

#Filter codes from the codelists 
dep_anx_psy_tx_usage = snomed_usage |> 
  filter(snomed_code %in% depr_anx_psy_tx_cod$code)
```

# Preliminary Objectives 

Preliminary objectives of the analysis include to:
- Describe the total usage of anxiety disorders and psychological therapy codes over the study period 
- Identify the most frequently used psychological therapy codes in English Primary care 
- Examine pattern of code usage by:
  - Referral or signposting vs actual delivery of therapy
  - High-intensity vs low-intensity psychological therapies
  - Therapy indication, including specific codes for anxiety disorders, depression and cross-cutting codes


#Results 

```{r}
#| label: total-code-usage
#| echo: true
#| message: false
#| warning: false

#Obtain total code usage
dep_anx_tx_total_cod_usage = dep_anx_psy_tx_usage |>
group_by(end_date) |> 
    summarise(usage = sum(usage), .groups = "drop") |>
    mutate(ratio_usage = usage / sum(usage)) |>
    arrange(-usage)

```

```{r}
#| label: fig-total-trends
#| fig-cap: "Yearly counts of psychological therapy codes' from 2011/12 to 2023/24."
#| fig-height: 4
#| fig-width: 8
#| warning: false
 
fig_psy_tx_dep_anx_trends = dep_anx_tx_total_cod_usage |>
  ggplot(
      aes(
        y = usage,
        x = end_date,
      )
    ) +
    geom_line(alpha = 0.2) +
    geom_point() +
    labs(
      x = NULL,
      y = NULL,
      colour = NULL,
      caption = "Data from NHS England SNOMED Code Usage in Primary Care.",
      title = "Yearly counts of psychological therapy codes used in primary care"
    ) +
    scale_x_date(
      date_breaks = "1 year",
      labels = label_date_short()
    ) +
    scale_y_continuous(
      labels = label_number(accuracy = 1),
      limits = c(0, NA)
    ) +
    theme_bw() +
    theme(legend.position = "bottom") +
  scale_color_viridis_d()
fig_psy_tx_dep_anx_trends

ggsave(
  filename = here("output", "figures", "fig_psy_tx_dep_anx_total_trends.png"),
  fig_psy_tx_dep_anx_trends,
  height = 4,
  width = 8
)

```













